Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-08 Manuscript/manuscript.tex
Update to the Stoichiometry of binding section -- mainly corrections and comments on what to do.

Binding stoichiometry of \ce{Ca^{2+}} and POPC was 
thoroughly studied in the experimental work~\cite{altenbach84}.
Several binding models were proposed and tested of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca^{2+}} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
In a later work~\cite{macdonald87}, 
a Langmuir adsorption model (i.e. stoichiometry 1~POPC:1~\ce{Ca^{2+}}) 
was found to provide as good fit as ternary complex model, 
when only low concentrations of \ce{CaCl2} are considered.

Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
Simulations, however, allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca^{2+}}:$n\times$POPC clusters
In Figure~\ref{fig:cacl_complexes} we see that ternary complex is indeed 
the most probable binding mode of calcium at 285~mM concentration. 
Apart from this complex, we also find complexes with 1~and 3~lipids
occuring with only a slightly lower but similar probability.
The fractions of \ce{Ca^{2+}}:$n\times$POPC complexes at 285~mM concetration 
are then in order:
$42\%$ for two lipids,
$30\%$ for one lipid, and
$28\%$ for three lipids.
The symmetry of such a distribution --  
i.e. almost equal probabilities of complexes with 1 or 3 lipids 
that behave in the total average picture as complexes with 2 lipids -- 
provides clues why ternary complex binding model 
fits both simulation and experimental results relatively well, 
although it is apparently incorrect.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/number_bound_lipids_p_cation_Ecc-lipids.eps} \\
  \caption{\label{fig:cacl_complexes}
      Histogram of relative probabilities of existence of complexes
      of \ce{Ca^{2+}} and $n$ POPC lipids. 
  }
\end{figure}

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
Based on the probability isodensity contours (see Fig.~\ref{fig:volmaps}), 
we estimate that the largest contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Although the isodensity plots are relatively easy to interpret,
the contours shown in Fig.~\ref{fig:volmaps} cannot conclusively tell on the details
of \ce{Ca^{2+}} binding to any of the two carbonyl moieties
but their apparently lower affinity compared to phosphate oxygens.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally determined to be lower than $10\,\mu\mathrm{s}$~\cite{carrington67}. 
Recent theoretical work with long enough simulations estimates this time
in the order of $1\,\mu\mathrm{s}$~\cite{catte16}. 
This is in contrast to our model, ECC-lipids, 
which gives a mean residence time in the order of $10\,\mathrm{ns}$, 
i.e. at least two orders of magnitude lower than previous estimates.
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent.

>>>>

This section is rough and will likely require editing. 
Data -- up to noted exceptions -- shall be all there, however.

Binding stoichiometry of \ce{Ca^{2+}} and POPC was 
thoroughly studied in the experimental work~\cite{altenbach84},
in which the head group order parameter changes to cation binding are determined.
Several binding models were proposed and tested of which only one,
ternary complex binding model, provided a good fit of the experimental observations.

Simulations allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca^{2+}}:$n\times$POPC clusters
by evaluating contacts between cations and lipids with a cut off radius $0.3\,\mathrm{nm}$. 
In Figure~\ref{fig:cacl_complexes} we see that ternary complex is indeed 
the most probable binding mode of calcium at 285~mM concentration. 
Apart from this complex, we also find complexes with 1~and 3~lipids
occuring with only a slightly lower but similar probability.
The fractions of \ce{Ca^{2+}}:$n\times$POPC complexes at 285~mM concetration 
are then in order:
$42\%$ for two lipids,
$30\%$ for one lipid, and
$28\%$ for three lipids.

Several binding models were proposed and tested~\cite{altenbach84} of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca^{2+}} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
In a later work~\cite{macdonald87}, 
a Langmuir adsorption model (i.e. stoichiometry 1~POPC:1~\ce{Ca^{2+}}) 
was found to provide as good fit as ternary complex model, 
when only low concentrations of \ce{CaCl2} are considered.
Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
The symmetry of the distribution of complexes from simulation --  
i.e. almost equal probabilities of complexes with 1 or 3 lipids 
that behave in the total average picture as complexes with 2 lipids -- 
provides clues why ternary complex binding model 
fits both simulation and experimental results relatively well, 
although it is apparently incorrect.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/number_bound_lipids_p_cation_Ecc-lipids.eps} \\
  \caption{\label{fig:cacl_complexes}
      Histogram of relative probabilities of existence of complexes
      of \ce{Ca^{2+}} and $n$ POPC lipids. 
  }
  \todo{Change this figure so that it contains the relative probabilities of finding the particular clusters.}
\end{figure}

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
\todo{Add a simple analysis using number of contacts.}
Based on the probability isodensity contours (see Fig.~\ref{fig:volmaps}), 
we estimate that the largest contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Although the isodensity plots are relatively easy to interpret,
the contours shown in Fig.~\ref{fig:volmaps} cannot conclusively tell on the details
of \ce{Ca^{2+}} binding to any of the two carbonyl moieties
but their apparently lower affinity compared to phosphate oxygens.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From recent theoretical work with long enough simulations this time can be roughly estimated
in the order of $1$--$10\,\mu\mathrm{s}$~\cite{Javanainen2017}. 
This is in contrast to our model, ECC-lipids, 
which gives a mean residence time in the order of $10\,\mathrm{ns}$, 
\todo{evaluate this number, mean residence time, accurately based on the contacts data.}
i.e. at least two orders of magnitude lower than previous estimates.
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent.
